The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-09-20 paper.qmd
Updated comments/responses

This does not mean that visualizing raw data will always prevent insignificant signal from getting through. 
@Chowdhury illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the line-up protocol was implemented. 
However, just showing the data it is simple but effective option for uncertainty visualisation that that seems to be largely overlooked. 
<!-- Does showing the data actually show uncertainty, or does it just show variance? I'm thinking of inferential statistics here - you can show a sample, but it doesn't necessarily show you anything about the uncertainty of the population-based inferential statistics, visually speaking. Some of the dotmap style visualizations you're thinking of are actually census visualizations - there isn't any inferential uncertainty, but there is variance. I'm not sure you're adequately separating these concepts here, even if I generally agree with the broad points you're making.  -->
<!--    
- I think the point I’m trying to make here is not that the raw data shows uncertainty by showing variation, but rather that it shows uncertainty by allowing people to see violations to the assumptions they might have made when mentally calculating an inferrential statistic
  - e.g. if i show the raw data and the areas are unevenly spaced or there are a lot of missing observations in some areas.
- I agree that the variability in the raw data is not related to the variability in the inferential statistic even though this section certainly implies that. I need to fix that.
- The connection with the pixel map was supposed to be that it would theoretically give a sense of the shape of the distribution. When I said “where the variability is coming from” I was trying to refer to that shape, but that doesn’t really translate from what I’ve written and the connecton between the pixel map and showing the raw data isn’t really there. 
  - I think the logic line I was going for was
    Pixel map shows collection of hypothetical outcomes of sampling distribution 
    => These outcomes let us see the shape of the distribution 
    => When seeing the shape of the distribution, we may become aware of errors in our assumptions, (e.g. such as normality) 
    => If we want to be aware of where our uncertainty is coming from (e.g. underlying variability or assumption violations) then the raw data is what best shows this
- I did consider including a visualisation that shows the data I used for the pixel map as a final map visualisation to better separate these concepts, but I didn’t technically use any real data (since I just made the normality assumption which is excatly what this map should highlight)
  - It might be better to change the statistic and give the sampling distribution in each area a different shape to make the pixel map benefit clearer.
  - I just want to make sure this is an accurate description of the issue you were highlighting before I fix it up this section (also I think it will take me a bit to figure out how to fix it) 
-->
While the pixel map does not literally show the data, it does imitate showing the data and give us details on where the variability is coming from. 
In order to uncerstand *why* pixel maps seem to be successful at signal suppression at first glace would likely require a more thorough discussion than what what we can offer here.

@@

I agree with your point about the accuracy in display vs perception but I don’t see its connection to the greater point about trust and I’m not 100% sure it fits in this section. Accuracy in the visual form (e.g. correcting for sine illusion) is still a pursuit of accuracy, just accuracy in the visual statistic.  Like, if we use Frequentist or Bayesian statistics, they also have different inputs, processes and outputs and they both give us different estimates. It is also common to adjust those statistics to correct for a bias in the calculation (e.g. we might weight the observations). I don’t really see the difference between these methods and editing a graphic to correct for the sine illusion. I don’t think any of these things conflict with the pursuit of scientific accuracy. I’m not against adding this as an additional comment (i.e. that accuracy in a visualisation may not be what the reader thinks it is) but I’m not certain it has any serious implications for the trust
-->
Setting trust as the variable of interest implicitly encourages statisticians to set trust and as the primary goal of communication. 
Evaluating visualisations on trust conflates trust and transparency and ultimately discourages signal-suppression. 
We can see this effect pop up in the results of evaluation studies. 
For example, @Zhao2023 found that participants were more trusting of model estimates with low uncertainty, but this effect did not carry over to estimates with high uncertainty. 
Despite decreased trust being a desired outcome of signal suppression, the authors discussion implied this result was not desired. 
This perspective ends up extending to visualisation authors as well.
@Hullman2020a found that authors simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy". 
This is a classic example of the negative impacts of placing direct importance on *trust* rather than *transparency*. 
In cases of high uncertainty, authors will opt to leave out uncertainty information because it decreases confidence in the authors conclusions. 
This is the end result of designing uncertainty visualisations for increased trust, rather than signal-suppression.

>>>>

This does not mean that visualizing raw data will always prevent insignificant signal from getting through. 
@Chowdhury illustrated how groups that appear linearly separable in a linear discriminant analysis (LDA) visualisation of the data can actually be the result of a LDA performed on too many variables, something that was not clear from the visualisation until the line-up protocol was implemented. 
However, just showing the data it is simple but effective option for uncertainty visualisation that that seems to be largely overlooked. 
<!-- Does showing the data actually show uncertainty, or does it just show variance? I'm thinking of inferential statistics here - you can show a sample, but it doesn't necessarily show you anything about the uncertainty of the population-based inferential statistics, visually speaking. Some of the dotmap style visualizations you're thinking of are actually census visualizations - there isn't any inferential uncertainty, but there is variance. I'm not sure you're adequately separating these concepts here, even if I generally agree with the broad points you're making.  -->
<!--    
- I think the point I’m trying to make here is not that the raw data shows uncertainty by showing variation, but rather that it shows uncertainty by allowing people to see violations to the assumptions they might have made when mentally calculating an inferrential statistic
  - e.g. if i show the raw data and the areas are unevenly spaced or there are a lot of missing observations in some areas.
- I agree that the variability in the raw data is not related to the variability in the inferential statistic even though this section certainly implies that. I need to fix that.
-->
<!-- Ok, but variability in the raw data is also related to variability in the inferential statistic, it's just that as n -> infinity, the variability in the inferential statistic goes to 0.  -->
<!--
- The connection with the pixel map was supposed to be that it would theoretically give a sense of the shape of the distribution. When I said “where the variability is coming from” I was trying to refer to that shape, but that doesn’t really translate from what I’ve written and the connecton between the pixel map and showing the raw data isn’t really there. 
  - I think the logic line I was going for was
    Pixel map shows collection of hypothetical outcomes of sampling distribution 
    => These outcomes let us see the shape of the distribution 
    => When seeing the shape of the distribution, we may become aware of errors in our assumptions, (e.g. such as normality) 
    => If we want to be aware of where our uncertainty is coming from (e.g. underlying variability or assumption violations) then the raw data is what best shows this
- I did consider including a visualisation that shows the data I used for the pixel map as a final map visualisation to better separate these concepts, but I didn’t technically use any real data (since I just made the normality assumption which is excatly what this map should highlight)
  - It might be better to change the statistic and give the sampling distribution in each area a different shape to make the pixel map benefit clearer.
  - I just want to make sure this is an accurate description of the issue you were highlighting before I fix it up this section (also I think it will take me a bit to figure out how to fix it) 
-->
<!-- Yes, this looks like you're thinking along the same lines as I was -->
While the pixel map does not literally show the data, it does imitate showing the data and give us details on where the variability is coming from. 
In order to uncerstand *why* pixel maps seem to be successful at signal suppression at first glace would likely require a more thorough discussion than what what we can offer here.

@@

I agree with your point about the accuracy in display vs perception but I don’t see its connection to the greater point about trust and I’m not 100% sure it fits in this section. Accuracy in the visual form (e.g. correcting for sine illusion) is still a pursuit of accuracy, just accuracy in the visual statistic.  Like, if we use Frequentist or Bayesian statistics, they also have different inputs, processes and outputs and they both give us different estimates. It is also common to adjust those statistics to correct for a bias in the calculation (e.g. we might weight the observations). I don’t really see the difference between these methods and editing a graphic to correct for the sine illusion. I don’t think any of these things conflict with the pursuit of scientific accuracy. I’m not against adding this as an additional comment (i.e. that accuracy in a visualisation may not be what the reader thinks it is) but I’m not certain it has any serious implications for the trust
-->
<!-- That's totally fair :) Just quibbling - not something you have to change unless someone else calls it out! -->
Setting trust as the variable of interest implicitly encourages statisticians to set trust and as the primary goal of communication. 
Evaluating visualisations on trust conflates trust and transparency and ultimately discourages signal-suppression. 
We can see this effect pop up in the results of evaluation studies. 
For example, @Zhao2023 found that participants were more trusting of model estimates with low uncertainty, but this effect did not carry over to estimates with high uncertainty. 
Despite decreased trust being a desired outcome of signal suppression, the authors discussion implied this result was not desired. 
This perspective ends up extending to visualisation authors as well.
@Hullman2020a found that authors simultaneously argued that failing to visualise uncertainty was akin to fraud, but also many avoided uncertainty visualisation because they didn't want their work to come across as "untrustworthy". 
This is a classic example of the negative impacts of placing direct importance on *trust* rather than *transparency*. 
In cases of high uncertainty, authors will opt to leave out uncertainty information because it decreases confidence in the authors conclusions. 
This is the end result of designing uncertainty visualisations for increased trust, rather than signal-suppression.
